Analysing the three-dimensional structure of {\em P. falciparum}

### 2017-07-09 doc/manuscript/paper.tex
Progress on the paper

% Title of paper
\title{Analysing the three-dimensional structure of P. falciparum}
\author{Nelle Varoquaux}

@@

\section{Introduction}
\label{sec1}

\section{Methods}
\label{sec2}

\subsection{Analysing contact count data}

- Normalization
- Quality assessment
- Enrichment of counts
- Anything else?
    - Did we do a compartment analysis?
- Can we extend it?
  - Fit-Hic ?

\subsection{Inferring the 3D structures}

@@

\subsubsection{Constrained multi-dimensional scaling}

@@

\subsubsection{Reproducing the results using a simpler model}

@@

\subsubsection{Structure stability}

@@

A number of methods have been proposed to accomplish such tasks, which I will
review in this section.

\subsubsection{Volume exclusion models}

@@

\subsection{Integrative analysis of gene expression and 3D structure using KernelCCA}

@@

\section{Results}
\label{sec3}

>>>>

% Title of paper
\title{Analysing the three-dimensional structure of {\em P. falciparum}}
\author{Nelle Varoquaux}

@@

\section{Introduction}
\label{sec:introduction}

With more than 217 millions cases nearly 429 000 deaths in 2015,  malaria
remains a major disease burden in tropical and subtropical countries and an
impediment to economic development. In Africa, where more than 90\% of the
cases and deaths occurs, the effect of malaria extend far beyond direct
measures of mortality: the disease is thought to cost more than US\$21
billion a year \citep{onwejkwe:do}.

In humans, malaria is caused by a unicellular protozoan {\em Plasmodium} of
the following species: {\em P. vivax}, {\em P. malariae}, {\em P. ovale}, {\em
P. knowelis}, and with 99\% of the deaths associated to it, by far the most
deadly {\em P. falciparum}. The parasite has a complex life cycle, with
multiple stage both in the human and mosquito host. The human host is infected
with sporozoites through the bite of a infected female  {\em Anopheles}
mosquito. The sporozoites quickly migrate to the liver, where they start a two
weeks multiplication process. They are then released as merozoites in the
bloodstream, and infect red blood cells. The parasites then start their
"erythrocytic" stage, via another round of replications in red blood cells,
through an unusual process of cell division called schizogony. The parasite
first undergoes multiple rounds of nuclear replication and division  into 13
to 32 daughter cells, until the red blood cell bursts and provokes the release
of merezoites and the infective cycle starts anew. This asexual replication
cycle is responsible for the symptoms and the complications of the disease:
anemia, tertian fever, \dots

While efficacious vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citet{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinating changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms . Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at the
epigenetic and post-translational levels
\citep{cui:chromatin-mediated, duffy:role, hoeijmakers:placing,
horrocks:control, deitsch:mechanisms}.

\fixme{introduction on HiC}

We discuss below the analysis methods used and developed for studying the 3D
architecture of {\em P. falciparum}'s genome.

\section{Profiling {\em P. falciparum}'s genome architecture during the
asexual cycle}
\label{sec:data}

- Difficulty of the protocol on the AT-rich genome of p. falciparum.
- Summary of the data

\section{Analyzing contact count data}

- Normalization
- Quality assessment
- Enrichment of counts
- power-law analysis
- 
- Anything else?
    - Did we do a compartment analysis?
- Can we extend it?
  - Fit-Hic ?

\section{Inferring the 3D structures}

@@

\subsection{Constrained multi-dimensional scaling}

@@

\subsection{Reproducing the results using a simpler model}

@@

\subsection{Structure stability}

@@

\section{3D gene set enrichment}

To assess whether groups of genes are colocolized in a 3D model,
\citet{ay:three-dimensional} leveraged a statistical method developped by
\citet{witten:XXX}, which requires labeling each pair of loci in two groups:
"close" or "far".

The authors used varying distance
thresholds (10\%, 20\% and 40\% of the nuclear diameter) to deem a locus pair
“close” and labeled all remaining
pairs in the set as “far”.

\fixme{Re-read Witten and Noble}

This approach dichotomize loci pairs into two groups, and check for the
enrichment of a label in one of the two groups. \citep{capurso:distance-based}
presents an approach that avoids this step, and instead directly estimates the
significance within the 3D model. Briefly, for a group $\mathcal{G}$, MPED
computes a test statitics:
\begin{equation*}
M = \underset{i,j \in \mathcal{G}| c_i != c_j}{\text{median}} d_{ij}\,,
\end{equation*}

where $d_{ij}$ is the euclidean distance between bead $i$ and bead $j$. The
null distribution is estimated empirically by resampling $10^5$ with
preservation of the chromosome structure. If the $M$ statistics is smaller
than the mean of the null distribution, it is compared to the lower tail of
the distribution and indicates co-localization. If the $M$ statistics is
larger than the mean of the null distribution, then it is compared to the
upper tail of the distribution, and indicates dispersion.

Applying MEPD to the Ring stage, \citet{capurso:distance-based} confirm that
centromers, telomers, VRSM genes (both overall, subtelomeric and internal)
colocalize.

\section{Volume exclusion models cannot explain }

@@

\section{Integrative analysis of gene expression and 3D structure using KernelCCA}

@@
